---
title: "p8105_hw3_yz4990"
author: "Yucheng Zhao"
date: "2024-10-12"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)

```


## Problem 2
```{r}
demo_df = read_csv("./data/nhanes_covar.csv", 
                   na = "NA", 
                   skip = 4) |> 
  janitor::clean_names() |> 
  filter(age >= 21) |> 
  drop_na() |> 
  mutate(sex = as.factor(sex),
    education = as.factor(education))


mims_df = read_csv("./data/nhanes_accel.csv", 
                   na = "NA",
                   ) |> 
  mutate(across(min1:min1440, as.double)) |> 
  janitor::clean_names() |> 
  pivot_longer(
    cols = min1:min1440,
    names_to = "interval_min", 
    names_prefix = "min", 
    values_to = "mims"
  )

acce_df = full_join(mims_df, demo_df, by = "seqn") |> 
  filter(age >= 21)

```

reader-friendly table of Number of Men and Women in Each Education Category
```{r}
gender_table = demo_df |> 
  select(education, sex) |> 
  group_by(education) |> 
  count(education, sex) |> 
  
  pivot_wider(
    names_from = "sex",
    values_from = "n"
  ) |> 
  group_by(education) |> 
  mutate(
    education = factor(education, 
                       levels = c(1, 2, 3), 
                       labels = c("Less than high school", 
                         "High school equivalent", 
                         "More than high school"))
  ) |> 
  rename(Male = "1", Female = "2", Education = "education")

knitr::kable(gender_table, caption = "Number of Men and Women in Each Education Category")
  
```
The table is reader-friendly and shows the number of males and females in each education category. 

Less than high school: There are 27 males and 28 females in this category. 

High school equivalent: There are 35 males and 23 females in this catgory. 

More than high school: There are 56 males and 59 females in this category.




```{r}
demo_df |> 
  select(age, sex, education) |> 
  ggplot(aes(x = age, fill = sex)) +
  geom_histogram(binwidth = 10, position = "dodge") +
  facet_grid(~ education, labeller = as_labeller(c(
    `1` = "Less than high school",
    `2` = "High school equivalent",
    `3` = "More than high school"
  ))) +
  scale_fill_discrete(labels = c("Male", "Female")) + 
  labs(
    title = "Age Distribution by Sex within Education Levels",
    x = "Age",
    y = "Count",
    fill = "Sex"
  ) +
  theme_minimal()



```



Density plot of age distribution by sex in each education category
```{r}
demo_df |> 
  select(sex, age, education) |> 
  ggplot(aes(x = age, fill = sex)) +
  geom_density(alpha = 0.4, adjust = .5) + 
  facet_grid(~ education, labeller = 
               as_labeller(c("1" = "Less than high school",
                             "2" = "High school equivalent",
                             "3" = "More than high school"))) +
  scale_fill_discrete(labels = c("Male", "Female")) + 
  labs(
    title = "Age Distribution by Sex within Education Categories",
    x = "Age",
    y = "Density",
    fill = "Sex"
  ) +
  theme_minimal() 

```
This density plot provides the age distributions for men and women in each education category. 

Less than high school: Both male and female in this category show peaks around age 50 and age 70. There is a significant amount of overlaps between male and female, indicating that males and females with a "Less than high school" education have a similar age distribution. 

High school equivalent: The density curve for male shows a peak around age 25 and 55, while the curve for female shows a peak around age 50 and a larger peak around age 80. There is less overlap, indicating there are more age differences between males and females with a "High school equivalent" education. It suggests that younger men are more likely to have a "High school equivalent" education than women, while older women are more likely to have this education than men. 

More than high school: The density curves for both male and female have peaks around age 30, indicating that younger individuals of both gender are more likely to have a higher education. The peak of the female curve is higher than male's, showing that young women are more prominent in this category. Both curves decrease sharply after age 65, indicating fewer older individuals with the "more than high school" education.  


aggregate across minutes to create a total activity variable for each participant
```{r}
total_df = acce_df |> 
  group_by(seqn) |> 
  summarize(total_activity = sum(mims)) |> 
  left_join(demo_df, by = "seqn")

```



plot total activities against age, separate panels for education, include trend line
```{r}
ggplot(total_df, aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.5) +  
  geom_smooth(se = FALSE) +  
  facet_grid(~ education, labeller = as_labeller(c(
    `1` = "Less than high school",
    `2` = "High school equivalent",
    `3` = "More than high school"
  ))) +
  labs(
    title = "Total Activity vs Age for Males and Females by Education Level",
    x = "Age",
    y = "Total Activity"
  ) +
  viridis::scale_color_viridis(name = "Sex", 
                               labels = c("Male", "Female"), 
                               discrete = TRUE) + 
  theme_minimal() + 
  theme(legend.position = "bottom")

```
The scatter plot provides total activity vs age for males and females by each education level. The plot compares men to women and includes a trend line to illustrate differences in each category. 

Less than high school: Women shows a higher total activity from age 21 to 40 but then surpassed by men after age 40. Men's total activity has a peak around age 60. Both men and women have a sharp decline in total activity after age 60. 

High school equivalent: Both men and women have a peak around age 40 but then declined sharply from age 40 t0 60. Women have higher total activity overall. 

More than high school: Both men and women have a more flat trend line, indicating a slower decline in total activity as age increases. Women have higher total activity overall. 

In general, The plot shows that higher education links with a more stable activity level through different ages. While total activity decreases with age, the rate of decreasing is lower for individuals with higher education.  Women with "less than high school" education are generally less active than men, while women with "high school equivalent" or "more than high school" education are more active than men. 


three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex
```{r}
hour_df = acce_df |> 
  mutate(
    hour = pmin(as.numeric(str_extract(interval_min, "\\d+")) %/% 60 + 1, 24)
    ) |> 
  group_by(seqn, hour, sex, education) |> 
  summarize(hour_activity = sum(mims, na.rm = TRUE), .groups = "drop")

ggplot(hour_df, aes(x = hour, y = hour_activity, color = sex)) +
  geom_point(alpha = 0.3) +  
  geom_smooth(se = FALSE) +  
  facet_grid(~ education, labeller = as_labeller(c(
    `1` = "Less than high school",
    `2` = "High school equivalent",
    `3` = "More than high school"
  ))) +
  labs(
    title = "24-Hour Activity Time Courses for Males and Females by Education Level",
    x = "Hour",
    y = "Hourly Activity"
  ) +
  viridis::scale_color_viridis(name = "Sex", 
                               labels = c("Male", "Female"), 
                               discrete = TRUE) + 
  theme_minimal() + 
  theme(legend.position = "bottom") 
  
```
Less than high school: Women and men in this panel have very similar trends in activity levels through the 24 hours. Both groups show a peak of activity around 12PM and steadily decline after 8PM.

High school equivalent: Both women and men in this panel have a slightly lower activity level compared to individuals with "less than high school" or "more than high school" education. Men in this group have a lower activity level than women. 

More than high school: Overall, women in this panel have a noticeably higher activity level than men, especially during midday and afternoon.

In general, Men and women in all education panels share a similar activity patterns. The activity levels reach their peak around 12PM and then sharply decreases from 8PM to midnight. 


## Problem 3
Files contain 1% of all rides with a total duration less than 4 hours in each of four months. Import, clean, and tidy these data, and describe the resulting dataset.
```{r}
jan_20 = read_csv("./data/Jan 2020 Citi.csv") |> 
  mutate(month = "january",
         year = 2020)

jan_24 = read_csv("./data/Jan 2024 Citi.csv") |> 
  mutate(month = "january",
         year = 2024)

july_20 = read_csv("./data/July 2020 Citi.csv") |> 
  mutate(month = "july",
         year = 2020)

july_24 = read_csv("./data/July 2024 Citi.csv") |> 
  mutate(month = "july",
         year = 2024)

cb_df = bind_rows(jan_20, july_20, jan_24, july_24) |> 
  pivot_longer(
    start_station_name:end_station_name,
    names_to = "start_end",
    values_to = "station_name"
  ) |> 
  mutate(
    start_end = str_replace(start_end, "start_station_name", "start"),
    start_end = str_replace(start_end, "end_station_name", "end")
  ) |> 
  select(ride_id, rideable_type, weekdays, duration, member_casual, start_end, station_name, month, year)
  
```
pivot_longer(
    start_station_name:end_station_name,
    names_to = "start_end",
    values_to = "station_name"
  ) |> 
  mutate(
    start_end = str_replace(start_end, "start_station_name", "start"),
    start_end = str_replace(start_end, "end_station_name", "end")
  ) |> 
  select(ride_id, rideable_type, weekdays, duration, member_casual, start_end, station_name, month, year)
  
  
  
  
  Since the 3 datasets are already in a standard format with no missing values, no additional data cleaning was required. Each dataset initially contains 7 variables, including "ride_id," "rideable_type," "weekdays," "duration," "start_station_name," "end_station_name," and "member_casual." These variables are consistent within each dataset. 

To distinguish the month and year of the rides, I added 2 new columns "month" and "year" to each dataset before combining. After merging the datasets for January and July 2020 and 2024, I changed the combined dataset into a tidy format. The "start_station_name" and "end_station_name" columns are pivoted into 2 new columns "start_end" and "station_name." The "start_end" indicates whether the station is a start or end station for the ride, while the "station_name" contains the name of the corresponding station. 


a reader-friendly table showing the total number of rides in each combination of year and month separating casual riders and Citi Bike members. Comment on these results.
```{r}
cb_df |> 
  group_by(year, month, member_casual) |> 
  summarise(total_rides = n_distinct(ride_id), .groups = "drop") |> 
  pivot_wider(
    names_from = member_casual,
    values_from = total_rides
  ) |> 
  knitr::kable(caption = "Total Number of Rides by Year, Month, and Rider Type")
```
The table above shows the total number of rides in each combination of year and month separating casual riders and Citi Bike members. 

In January 2020, there were 984 rides by casual riders and 11436 rides by members in the data. 

In July 2020, there were 5637 rides by casual riders and 15411 rides by members in the data. 

In January 2024, there were 2108 rides by casual riders and 16753 rides by members in the data. 

In July 2024, there were 10894 rides by casual riders and 36262 rides by members in the data. 


a table showing the 5 most popular starting stations for July 2024; include the number of rides originating from these stations
```{r}
cb_df |>
  filter(month == "july",
         year == 2024, 
         start_end == "start") |> 
  group_by(station_name) |> 
  summarize(number_rides = n_distinct(ride_id), .groups = "drop") |>
  arrange(desc(number_rides)) |> 
  slice(1:5) |> 
  knitr::kable(caption = "The 5 Most Popular Starting Stations for July 2024")
```
demo_df |> 
  select(age, sex, education) |> 
  ggplot(aes(x = age, fill = sex)) +
  geom_histogram(binwidth = 10, position = "dodge") +
  facet_grid(~ education, labeller = as_labeller(c(
    `1` = "Less than high school",
    `2` = "High school equivalent",
    `3` = "More than high school"
  ))) +
  scale_fill_discrete(labels = c("Male", "Female")) + 
  labs(
    title = "Age Distribution by Sex within Education Levels",
    x = "Age",
    y = "Count",
    fill = "Sex"
  ) +
  theme_minimal()


Make a plot to investigate the effects of day of the week, month, and year on median ride duration. This plot can include one or more panels, but should facilitate comparison across all variables of interest. Comment on your observations from this plot.
```{r}
cb_df |> 
  group_by(weekdays, month, year) |> 
  summarize(median_duration = median(duration), .groups = "drop") |> 
  ggplot(aes(x = weekdays, y = median_duration, fill = weekdays)) +
  geom_col(position = "dodge") +  
  facet_grid(year ~ month) +
  labs(
    title = "Median Ride Duration by Weekdays, Month, and Year",
    x = "Weekdays",
    y = "Median Ride Duration",
    fill = "Weekdays"
  )  +
  viridis::scale_fill_viridis(name = "Weekdays", 
                               discrete = TRUE) + 
  theme_minimal() + 
  theme(legend.position = "bottom") 
```




```{r}

```




```{r}

```





